### 2018-09-27 Chapters/Chapter2.tex
Squres lattice

Additionally, for $i=j=k=l$ the Pauli principle requires $\sigma' = \sigma$

@@

At half filling, in the $t << \text{U}$ limit we can apply perturbation theory to obtain the Heisenberg model from the Hubbard Hamiltonian \ref{Hubbard}. First notice that in the case $t = 0$ the ground state corresponds to having all sites single-occupied. There are two possible spin orientations for each sites, so there is a $2^M$ degeneracy. The ground state energy is $0$. The first exited states are obtained by moving one electron from one site to another, thus leaving one empty site and one double-occupied site. The energy of one of these states is $\text{U}$. Now, if we turn the kinetic term on as a perturbation, having two neighbors electrons with antiparallel spin will allow them to hop and reduce the kinetic energy, whereas while being parallel this hopping is not allowed and there is no energy reduction. Therefore, we can see that when turning the kinetic term on antiparallel alignment will be favored.

Let us write $\hat{H} = \hat{H_0} + \hat{H_t}$, $\hat{H_t} = -\sum_{\langle i,j \rangle, \sigma, \sigma'}(\delta_{\sigma, \sigma'} t + \boldsymbol{\Delta}_{ij} \boldsymbol{\sigma}_{\sigma, \sigma'})\hat{c}_{i \sigma}^\dagger \hat{c}_{j \sigma'}$. Let $\ket{\phi}$ and $\ket{\phi'}$ be state from the ground state, then we know $\bra{\phi'} \hat{H_0} \ket{\phi} = 0$ and in first order in $t$ $\bra{\phi'} \hat{H} \ket{\phi} = \bra{\phi'} \hat{H_t} \ket{\phi} = 0$ since $\hat{c}_{i \sigma}^\dagger \hat{c}_{j \sigma}$ moves an electron from site $j$ to $i$ thus leaving $i$ doubly occupied, therefore $\hat{H_t} \ket{\phi}$ has no superposition with $\phi'$. In second order we have:

@@

Which is the Heisenberg model for antiferromagnets with $J = \frac{2t^2}{\text{U}}$. Likewise, for $\hat{H}_D$:

>>>>

i.e. $\text{U} = \int d\bs{r} d\bs{r}' \phi(\bs{r})^* \phi(\bs{r}')^* \hat{v}(\bs{r}-\bs{r}') \phi(\bs{r}') \phi(\bs{r})$, where $\phi(\bs{r})$ is the Wannier function centered at the origin. Additionally, for $i=j=k=l$ the Pauli principle requires $\sigma' = \sigma$.

@@

At half filling, in the $t << \text{U}$ limit, which correspond to the strong coupling limit, we can apply perturbation theory to obtain the Heisenberg model from the Hubbard Hamiltonian \ref{Hubbard}. First notice that in the case $t = 0$ the ground state corresponds to having all sites single-occupied. There are two possible spin orientations for each sites, so there is a $2^M$ degeneracy. The ground state energy is $0$. The first exited states are obtained by moving one electron from one site to another, thus leaving one empty site and one double-occupied site. The energy of one of these states is $\text{U}$. Now, if we turn the kinetic term on as a perturbation, having two neighbors electrons with antiparallel spin will allow them to hop and reduce the kinetic energy, whereas while being parallel this hopping is not allowed and there is no energy reduction. Therefore, we can see that when turning the kinetic term on antiparallel alignment will be favored.

Let us write $\hat{H} = \hat{H}_0 + \hat{H}_t$, where $\hat{H}_0$ is the Hamiltonian for $t=0$ i.e. $\hat{H}_0 = \text{U} \sum_{i=1}^M \hat{n}_{i\uparrow}\hat{n}_{i\downarrow}$ and $\hat{H_t} = -\sum_{\langle i,j \rangle, \sigma, \sigma'}(\delta_{\sigma, \sigma'} t + \boldsymbol{\Delta}_{ij} \boldsymbol{\sigma}_{\sigma, \sigma'})\hat{c}_{i \sigma}^\dagger \hat{c}_{j \sigma'}$. Let $\ket{\phi}$ and $\ket{\phi'}$ be states from the ground state of $\hat{H}_0$, then we know $\bra{\phi'} \hat{H_0} \ket{\phi} = 0$ and in first order in $t$ $\bra{\phi'} \hat{H} \ket{\phi} = \bra{\phi'} \hat{H_t} \ket{\phi} = 0$ since $\hat{c}_{i \sigma}^\dagger \hat{c}_{j \sigma}$ moves an electron from site $j$ to $i$ thus leaving $i$ doubly occupied, therefore $\hat{H_t} \ket{\phi}$ has no superposition with $\phi'$. In second order we have:

@@

Which is the Heisenberg model for antiferromagnets with $J = \frac{2t^2}{\text{U}} > 0$. Likewise, for $\hat{H}_D$:

### 2018-09-27 Chapters/Chapter3.tex
minor changes
Squres lattice

\begin{equation}
\hat{H} = -t_0\sum_{\langle i,j \rangle, \sigma} \hat{c}_{i \sigma}^\dagger \hat{c}_{j \sigma} + U \sum_{i=1}^M \hat{n}_{i\uparrow}\hat{n}_{i\downarrow}
\end{equation}

Which is the Hamiltonian \ref{Hubbard} without DMI interaction. We can introduce a time dependent electric field in this Hamiltonian by adding a phase to the hopping term $t_{ij}(t) = t_0e^{ie\bs{R}_{ij}\bs{A}(t)}$, where $\bs{A}$ is the vector potential and $\bs{R}_{ij} = \bs{R}_i-\bs{R}_j$. This is known as Peierls substitution and it is derived in \ref{AP3A}. The electric field has the form $\bs{E}(t) = E_0 \sin (\omega t)\bs{\hat{x}}$ (one-dimensional lattice). We can therefore take $\bs{A}(t) = -\frac{E_0}{\omega}\cos(\omega t)\bs{\hat{x}} = -\bs{A}\cos(\omega t)$. Therefore the time dependent Hamiltonian is:

\begin{equation}
\hat{H}(t) = -\sum_{\langle i,j \rangle, \sigma} t_{ij}(t)\hat{c}_{i \sigma}^\dagger \hat{c}_{j \sigma} + U \sum_{i=1}^M \hat{n}_{i\uparrow}\hat{n}_{i\downarrow} = -\hat{T}(t) + U\hat{D}
\end{equation}

Where $\hat{D}$ is the double occupancy number operator and $\hat{T}(t) = \sum_{\langle i,j \rangle, \sigma} t_{ij}(t)\hat{c}_{i \sigma}^\dagger \hat{c}_{j \sigma}$ is the hopping operator. Since the time dependence in periodic we can expand $t_{ij}(t) = \sum_m t_{ij,m} e^{im \omega t}$, where

\begin{equation}
t_{ij,m} = t_0(-1)^mJ_m(\alpha_{ij})
\end{equation}

Where $\alpha_{ij} = e\bs{R}_{ij}\bs{A}$ and $J_m(x)$ is the \textit{m}th Bessel function, and correspondingly $\hat{T}(t) = \sum_m \hat{T}_m e^{im \omega t}$.

In order to derive the form of the effective Hamiltonian in the half filling system with $\frac{t}{U} << 1$ let us introduce a time dependent unitary transformation $\hat{U}(t) = e^{-i\hat{S}(t)}$. The transformed Hamiltonian is:

@@

We perform the unitary transformation perturbatively in the hopping parameter $t_0$. We expand $\hat{S}(t) = \sum_\nu \hat{S}^{(\nu)}(t)$ and $\hat{H}'(t) = \sum_\nu \hat{H}'^{(\nu)}(t)$. In order for the new Hamiltonian to be periodic we impose the unitary transformation to have the periodicity of the Hamiltonian, so that we can expand $\hat{S}^{(\nu)}(t) = \sum_m e^{im\omega t}\hat{S}^{(\nu)}_m$. Additionally we can impose the transformed Hamiltonian to be block diagonal in the doublon number $d$. With these conditions the unitary transformation can be uniquely determined if we impose that $\hat{S}(t)$ does not contain block-diagonal terms, i.e. we can write:

\begin{equation}
\hat{S}^{(\nu)}(t) = \sum_{d \neq 0} \sum_m \hat{S}^{(\nu)}_{d,m} e^{im\omega t}
\end{equation}

@@

\begin{equation}
\hat{T}(t) = \sum_m (\hat{T}_{-1,m}+\hat{T}_{0,m}+\hat{T}_{1,m})e^{im\omega t}
\end{equation}

Where $\hat{T}_d(t)$ changes the double occupancy number by $d$, for example $\hat{T}_1(t) = \hat{P}_1 \hat{T}(t) \hat{P}_0$. We can write the transformed Hamiltonian as:

@@

To zeroth order we have $\hat{H}'^{(0)}= U\hat{D}$ and $i\hat{S}^{(0)}(t)=0$. In first order we obtain:

@@

\begin{align}
J_{ij}(t) &= \sum_{mn} \left( \frac{t_{ji,(m-n)} t_{ij,n}}{U+n\omega} + \frac{t_{ji,n} t_{ij,(m-n)}}{U-n\omega} \right) e^{im\omega t} \nonumber \\
&= \sum_{mn} (-1)^m t_0^2 \left( \frac{J_{m-n}(\alpha_{ji})J_n(\alpha_{ij})}{U+n\omega} +\frac{J_n(\alpha_{ji})J_{(m-n)}(\alpha_{ij})}{U-n\omega} \right)
\end{align}

In the case $\omega >> U$ we can limit to $m=0$ (time independent part) and $n=0$ or $n=\pm1$ (small denominators), additionally, if the electric field is small we can use $J_n(x) \approx x^n$. With these approximations:

\begin{equation}
J_{ij} = 2 t_0^2 \alpha_{ij}^2 \left( \frac{1}{U-\omega} - \frac{1}{U} + \frac{1}{U+\omega} \right)
\end{equation}

@@

\begin{equation}
t_{ij}(t) = \bra{\psi'_{\bs{R_i}}} \hat{H}'(\bs{r}) \ket{\psi'_{\bs{R_j}}} = e^{i\frac{e}{\hbar}\int_{\bs{R}_j}^{\bs{R}_i} d\bs{r}'\bs{A}(\bs{r}',t)} \bra{\psi_{\bs{R_i}}} \hat{H}(\bs{r}) \ket{\psi_{\bs{R_j}}}
\end{equation}

@@

>>>>

\begin{equation}
\hat{H} = -t_0\sum_{\langle i,j \rangle, \sigma} \hat{c}_{i \sigma}^\dagger \hat{c}_{j \sigma} + \text{U} \sum_{i=1}^M \hat{n}_{i\uparrow}\hat{n}_{i\downarrow}
\end{equation}

Which is the Hamiltonian \ref{Hubbard} without DMI interaction. We can introduce a time dependent electric field in this Hamiltonian by adding a phase to the hopping term $t_{ij}(t) = t_0e^{ie\bs{R}_{ij}\bs{A}(t)}$, where $\bs{A}$ is the vector potential and $\bs{R}_{ij} = \bs{R}_i-\bs{R}_j$. This is known as Peierls substitution and it is derived in \ref{AP3A}. We write the electric field as $\bs{E}(t) = \frac{1}{2}(\vec{E}e^{-i\omega t}+\vec{E}^*e^{i\omega t})$, where $\vec{E} = E_0\hat{e}$ and $\hat{e} = \frac{1}{\sqrt{1+\lambda_{POL}^2}}(\hat{e}_x+i\lambda_{POL}\hat{e}_y)$ is the polarization vector and $\lambda_{POL} = 0, \pm 1$ for plane polarized, right handed and left handed circular polarized field respectively. The vector potential takes the form $\bs{A}(t) = \frac{1}{2}(\vec{A}e^{-i\omega t} + \vec{A}^* e^{i\omega t})$, with $\vec{A} = \frac{iE_0}{\omega}\hat{e}$. The time dependent Hamiltonian is:

\begin{equation}
\hat{H}(t) = -\sum_{\langle i,j \rangle, \sigma} t_{ij}(t)\hat{c}_{i \sigma}^\dagger \hat{c}_{j \sigma} + \text{U} \sum_{i=1}^M \hat{n}_{i\uparrow}\hat{n}_{i\downarrow} = -\hat{T}(t) + \text{U}\hat{D}
\end{equation}

Where $\hat{D}$ is the double occupancy number operator and $\hat{T}(t) = \sum_{\langle i,j \rangle, \sigma} t_{ij}(t)\hat{c}_{i \sigma}^\dagger \hat{c}_{j \sigma}$ is the hopping operator. Let $e\bs{R}_{ij}\vec{A} = \alpha_{ij} e^{i \theta_{ij}}$, with $\alpha_{ij}$ a real number, then we can apply the Jacobi–Anger expansion:

\begin{align*}
t_{ij}(t) = t_0e^{ie\bs{R}_{ij}\bs{A}(t)} = t_0e^{i\alpha_{ij} \cos(\omega t - \theta_{ij})} = t_0 \sum_m e^{i(\frac{\pi}{2}-\theta_{ij})m} J_m(\alpha_{ij}) e^{im\omega t} = \sum t_{ij,m} e^{im\omega t}
\end{align*}

Where $t_{ij,m} = t_0 e^{i(\frac{\pi}{2}-\theta_{ij})m} J_m(\alpha_{ij})$ and $J_m(x)$ is the \textit{m}th Bessel function. Correspondingly $\hat{T}(t) = \sum_m \hat{T}_m e^{im \omega t}$.

In order to derive the form of the effective Hamiltonian in the half filling system with $t << \text{U}$ let us introduce a time dependent unitary transformation $\hat{U}(t) = e^{-i\hat{S}(t)}$. The transformed Hamiltonian is:

@@

We perform the unitary transformation perturbatively in the hopping parameter $t_0$, to that end let us write $\hat{T}(t) = \eta \hat{T}(t)$, where $\eta$ will play the role of a bookkeeping parameter in the perturbative expansion. We expand $\hat{S}(t) = \sum_\nu \eta^\nu \hat{S}^{(\nu)}(t)$ and $\hat{H}'(t) = \sum_\nu \eta^\nu \hat{H}'^{(\nu)}(t)$. In order for the new Hamiltonian to be periodic we impose the unitary transformation to have the periodicity of the Hamiltonian, so that we can expand $\hat{S}^{(\nu)}(t) = \sum_m e^{im\omega t}\hat{S}^{(\nu)}_m$. Additionally we can impose the transformed Hamiltonian to be block diagonal in the doublon number $d$. With these conditions the unitary transformation can be uniquely determined if we impose that $\hat{S}(t)$ does not contain block-diagonal terms, i.e. we can write:

\begin{equation}
\hat{S}^{(\nu)}(t) = \sum_{d \neq 0} \sum_m \eta^\nu \hat{S}^{(\nu)}_{d,m} e^{im\omega t}
\end{equation}

@@

\begin{equation}
\hat{T}(t) = \eta \sum_m (\hat{T}_{-1,m}+\hat{T}_{0,m}+\hat{T}_{1,m})e^{im\omega t}
\end{equation}

Where $\hat{T}_d(t)$ changes the double occupancy number by $d$, for example $\hat{T}_1(t) = \hat{P}_1 \hat{T}(t) \hat{P}_0$. Using the Baker–Campbell–Hausdorff expansion we can write the transformed Hamiltonian as:

@@

Now, starting from order zero and for each order we will determine $i\hat{S}^{(\nu)}(t)$ so that $\hat{H}'^{(\nu)}(t)$ is diagonal in the doublon number. To zeroth order in $\eta$ we have $\hat{H}'^{(0)}= U\hat{D}$ and $i\hat{S}^{(0)}(t)=0$. In first order we obtain:

@@

\begin{align}
J_{ij}(t) &= \sum_{mn} 2\left( \frac{t_{ji,(m-n)} t_{ij,n}}{U+n\omega} + \frac{t_{ji,n} t_{ij,(m-n)}}{U-n\omega} \right) e^{im\omega t} \nonumber \\
&= \sum_{mn} 2\left( \frac{t_{ji,(m-n)} t_{ij,n} + t_{ji,-n} t_{ij,(m+n)}}{U+n\omega} \right) e^{im\omega t} \label{Jij1}
\end{align}

Now, $t_{ij,m} = t_0 e^{i(\frac{\pi}{2}-\theta_{ij})m} J_m(\alpha_{ij})$, using that $\theta_{ji} = \theta_{ij}$ and $\alpha_{ji} = -\alpha_{ij}$ and the properties $J_{-m}(x) = (-1)^mJ_m(x)$ and $J_m(-x) = (-1)^mJ_m(x)$ we can write \ref{Jij1} as:

\begin{equation}
J_{ij}(t) = \sum_{mn} 2t_0^2 e^{i(\frac{\pi}{2}-\theta_{ij})m}\left(\frac{J_{n-m}(\alpha_{ij})J_{n}(\alpha_{ij})+J_{n}(\alpha_{ij})J_{n+m}(\alpha_{ij})}{U+n\omega} \right) e^{im\omega t}
\end{equation}

After time average this reduces to the $m=0$ term:

\begin{equation}
\label{Jij2}
J_{ij}(t) = \sum_{n} 4t_0^2 \frac{J_n(\alpha_{ij})^2}{U+n\omega}
\end{equation}

Now, $\alpha_{ij} = \text{Re}(e\bs{R}_{ij}\vec{A})$. For a square lattices aligned with the coordinate system so that $\vec{a}_1=\hat{e}_x$ and $\vec{a}_2=\hat{e}_y$, we will have $\bs{R}_{ij} = \pm a\hat{e}_x,\pm a\hat{e}_y$, where $a$ is the lattice constant. Using $\vec{A}=\frac{iE_0}{\omega\sqrt{1+\lambda^2}}(\hat{e}_x+i\lambda\hat{e}_y)$ we have:

\begin{equation}
\alpha_{ij} = \begin{cases}
             \pm \frac{eaE_0}{\omega \sqrt{1+\lambda^2}} = \pm \frac{\mathcal{E}}{\sqrt{1+\lambda^2}},  & \text{for } \bs{R}_{ij} = \pm \hat{e}_x \\
             \pm \lambda\frac{eaE_0}{\omega \sqrt{1+\lambda^2}} = \pm \frac{\lambda \mathcal{E}}{\sqrt{1+\lambda^2}},  & \text{for } \bs{R}_{ij} = \pm \hat{e}_y
       \end{cases} \quad
\end{equation}

Where $\mathcal{E} = \frac{eaE_0}{\omega}$. For $\omega>>U$ we can take only the three smallest values of $n$, i.e. $n=0, \pm 1$ in \ref{Jij2} as an approximation. We can also use that $\alpha_{ij} << 1$ since $\omega$ appears in the denominator and so $J_n(x) \approx x^n \text{ for } n>0 \text{ and } x << 1$:

\begin{equation}
J_{ij} = 4t_0^2 \left(\frac{\alpha_{ij}^2}{U+\omega} +\frac{1}{U} +\frac{\alpha_{ij}^2}{U-\omega} \right) = J_{ij}^0 + 4t_0^2 \alpha_{ij}^2 \left( \frac{1}{U+\omega} + \frac{1}{U-\omega} \right)
\end{equation}

For plane polarized light ($\lambda = 0$) the exchange interaction only changes in the direction of the polarization, that is:

\begin{equation}
J_{ij}^{PP} = \begin{cases}
		J_{ij}^0 + 4t_0^2 \mathcal{E}^2 \left( \frac{1}{U+\omega} + \frac{1}{U-\omega} \right) & \text{for } \bs{R}_{ij} = \pm \hat{e}_x \\
J_{ij}^0 & \text{for } \bs{R}_{ij} = \pm \hat{e}_y
\end{cases} \quad 
\end{equation}

For circular polarized light ($\lambda=\pm1$) in this approximation the exchange interaction changes in the same way in all the directions:

\begin{equation}
J_{ij}^{CP} = J_{ij}^0 + 2t_0^2 \mathcal{E}^2 \left( \frac{1}{U+\omega} + \frac{1}{U-\omega} \right)
\end{equation}

@@

\begin{equation}
\label{PeierlsMtxElem}
t_{ij}(t) = \bra{\psi'_{\bs{R_i}}} \hat{H}'(\bs{r}) \ket{\psi'_{\bs{R_j}}} = e^{i\frac{e}{\hbar}\int_{\bs{R}_j}^{\bs{R}_i} d\bs{r}'\bs{A}(\bs{r}',t)} \bra{\psi_{\bs{R_i}}} \hat{H}(\bs{r}) \ket{\psi_{\bs{R_j}}}
\end{equation}

@@

Notice that this substitution also applies if we add a Rashba Spin-Orbit interaction term to \ref{HamEMF}, i.e., if $\hat{H}'(\bs{r}) = \frac{(\bs{p}-e\bs{A})^2}{2m} +U(\bs{r}) + \alpha_R \hat{z}(\hat{\bs{\sigma}} \times (\bs{p}-e\bs{A}))$, then

\begin{align*}
  \alpha_R \hat{z}(\hat{\bs{\sigma}} \times (\bs{p}-e\bs{A})) \psi'_{\bs{R}} &= \alpha_R(\hat{\sigma_x}(\hat{p}_y-e\bs{A}) - \hat{\sigma_y}(\hat{p}_x-e\bs{A})) e^{i\frac{e}{\hbar}\int_{\bs{R}}^{\bs{r}}d\bs{r}'\bs{A}(\bs{r}',t)} \psi_{\bs{R}} = \\
  &= e^{i\frac{e}{\hbar}\int_{\bs{R}}^{\bs{r}}d\bs{r}'\bs{A}(\bs{r}',t)} \alpha_R \hat{z}\hat{\bs{\sigma}} \times \bs{p} \psi_{\bs{R}}
\end{align*}

Therefore relation \ref{PeierlsMtxElem} still holds.
